Lossy compression of DNA sequencing quality data

### 2015-04-09 manuscript/q-compression.tex
Added compression results.

\begin{figure*}[!tb]%figure2
%\centerline{\includegraphics[width=3.65in]{mse_frag.png}}
\centerline{\includegraphics[width=7in]{compression_results.pdf}}
\caption{Mean squared error versus bits/base-call for different
  compression methods applied to the fragment library from
  \textit{Rhodobacter sphaeroides} 2.4.1 data set. G/B: 2-bin
  encoding. P32 and P256: Quality profile compression with 32 and 256
  profiles. R0-R5: Polynomial regression-based compression using 0-5
  degrees. Dashed line corresponds to lossless bzip compression of
  quality values. \em}\label{fig:mse_vs_bpbp_frag}
\end{figure*}

@@

@@

We compare the \textsc{mse} versus bits/base-call of the fragment and
short-jump libraries of the \textit{Rhodobacter sphaeroides} 2.4.1
data set (Figures \ref{fig:mse_vs_bpbp_frag} and
\ref{fig:mse_vs_bpbp_jump}). Storing the uncompressed quality values
requires 1 byte per base-call because they are stored in
\textsc{ascii} format; however, lossless compression of the values
with BZip2 2.112 bits/base-call. 0-order polynomial regression and
32-profile encodings have the lowest bits/base-call. 5th-order
polynomial regression has the highest bits/base-call, but have among
the lowest \textsc{mse}. As the order of the polynomial increases, the
bits/base-call increase and the \textsc{mse} decreases at an
exponential rate.

>>>>

\begin{figure*}[!tb]%figure2
%\centerline{\includegraphics[width=3.65in]{mse_frag.png}}
\centerline{\includegraphics[width=7in]{compression_results.pdf}}
\caption{Mean squared error versus bits/base-call for different
  compression methods applied to the
  \textit{Rhodobacter sphaeroides} 2.4.1, and \textit{Homo sapiens} chromosome 14 fragment libraries, and \textit{Escherichia coli} str. K-12 MG1655, and \textit{Mus musculus} data sets. 2B: 2-bin
  encoding. P64, P128, and P256: Quality profile compression with 64, 128, and 256
  profiles. R0-R7: Polynomial regression-based compression using 0-7
  degrees. Dashed line corresponds to lossless Bzip2 compression of
  quality values. \em}\label{fig:mse_vs_bpbp}
\end{figure*}

@@

\textit{Mus musculus} data was downloaded from \url{http://trace.ddbj.nig.ac.jp/DRASearch/run?acc=SRR032209}, which consisted of 18,828,274 reads of length 36.

@@

We compare the \textsc{mse} versus bits/base-call of the \textit{Rhodobacter sphaeroides} 2.4.1, \textit{Homo sapiens} chromosome 14, \textit{Escherichia coli} str. K-12 MG1655, and \textit{Mus musculus} data sets (Figure \ref{fig:mse_vs_bpbp}). We only include the fragment libraries for the \textit{Rhodobacter sphaeroides} 2.4.1, and \textit{Homo sapiens} chromosome 14 data sets, but the additional short-jump library results are available in the supplementary (TODO:TABLE). Storing the uncompressed quality values
requires 1 byte per base-call because they are stored in
\textsc{ascii} format and is denoted by the dotted black line in the figure.
The lossless compression of each data set using BZip2 ranges from 2.19 - 3.10 bits/base-call and is denoted by the colored dotted lines.
The compression methods tend to cluster together across the different data sets.
Across all data sets, the 0-order polynomial regression, profile encodings, and QualComp have the lowest bits/base-call.

QualComp with the rate parameter set to 100 bits/read has the lowest \textsc{mse}, but requires 10-15x more storage than the profile encoding methods for only a 2-3x reduction in \textsc{mse}.
When QualComp's rate parameter is set to match the profile encoding methods, QualComp performs slightly worse in terms of \textsc{mse}.
In the \textit{Rhodobacter sphaeroides} 2.4.1 fragment library, QualComp with rate 10 bits/read (0.099 bits/bp) has a \textsc{mse} of 17.29.  Using 256-profile encoding requires less storage (0.079 bits/bp) and has a lower \textsc{mse} (11.85).

As the order of the polynomial increases, the
bits/base-call increase and the \textsc{mse} decreases at an
exponential rate.
The 7th-order polynomial regression has the highest bits/base-call and in the \textit{Mus musculus} data set, and requires more storage than the \textsc{ascii} original quality values.  A 7th-order polynomial requires storing eight floating point values, resulting in 32 bytes per quality sequence.  The read length of the \textit{Mus musculus} data set is only 26, so the 7th-order polynomial regression is storing six more bytes than necessary to losslessly encode the quality data.
